$p$-adic precision

### 2014-02-04 padicprec.tex
First tentative for Section 3.4. Comments are welcome.

@@

\todofor{Xavier}{
Explain here how to build a software that tracks precision automatically 
and sharply.
}

\subsubsection*{Working precision}

@@

In this Section, we illustrate the theory developed in this paper by 
giving examples and applications. In \S \ref{ssec:differentials}, we 
compute the differential of a bunch of usual operations involving 
$p$-adic objects of different nature. Very often, we will observe that 
this computation is easy even if the underlying operation is 
sophisticated (\emph{e.g.} Gr\"obner basis).
In \S \ref{ssec:SOMOS-solution}, we study the SOMOS 4 sequence 
introduced in \S \ref{ssec:stepbystep} and, making a crucial use of 
Lemma \ref{lem:main} and Proposition \ref{prop:}, we design a quite 
\emph{stable} algorithm for computing it.

@@

Another usual operations on polynomials is the factorization. Suppose 
that we are given a monic polynomial $P \in K[X]$ written as a product 
$P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
lemma implies that there exists a small neighbourhood $\mathcal U$ of 
$P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
defined on the subset of $\mathcal U$ consisting of monic polynomials.
We want to differentiate $f$ at $P_0$. For this, we differentiate the 
equality $P = A B$ around $P_0$, obtaining 
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

@@

Other important tools related to matrices are usual factorizations as, 
for example, LU factorization. In what follows, let us agree to define a 
LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
Such a writing exists and is unique provided that all principal minors 
of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
defined over the Zariski-open set of matrices satisfying the above 
condition. In order to differentiate it, we differentiate the relation 
$M = LU$ and rewrite the result as follows:
$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
We remark that in the right hand side of the above formula, the first
summand is lower triangular and all its diagonal entries vanish whereas 
the second summand is upper triangular. Hence in order to compute $dL$
and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
left and $U$ on the right respectively. 
The above discussion extends almost \emph{verbatim} to LUP 
factorization; the only difference is that LUP factorizations are not 
unique but they are on a small neighbourhood of $M$ if we fix the matrix 
$P$.

@@

Let us now consider the open subset of $M_{n,n-d}(K)$ consisting of 
matrices of full rank. The left kernel defines a mapping $\text{LK} : 
M_{n,n-d}(K) \to \Grass(d,n)$. Let us prove that it is differentiable 
and compute its differential around some point $M \in M_{n,n-d}(K)$. Of 
course, there exists a neighbourhood $\mathcal U$ of $M$ whose image is 
entirely contained in a given chart $U_P$. We assume for simplicity that 
$P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
our chart to the application that takes $M$ to the unique matrix $N \in
M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 
= 0$. The implicit function theorem then implies that $\text{LK}$ is
differentiable. Furthermore, its differential satisfies the relation
$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
from which we can compute $dN$ by projecting on the $(n-d)$ last
columns.

@@

The lesson from this study is the following. If we have to compute the 
kernel of a matrix $M$ over $K$ known with finite precision, \emph{a 
priori} we first need to determine what is the rank of $M$. However, the 
rank function is not continuous --- it is only lower semi-continuous --- 
and hence cannot be determined from sure from $M$ (expect if $M$ has 
full rank). We then need to make an extra hypothesis and the more natural 
one is arguably to assume that $M$ has actually the minimal rank that it 
can have. With this convention, for instance, the kernel of the zero 
matrix $M \in M_n(K)$ \emph{known with finite precision} is always the 
whole space although there exist invertible matrices in any 
neighbourhood of $M$.

@@

\ref{prop:} to $\varphi_{i-1}$ and $\varphi_i$ (at the point $(u_0,
u_1, u_2, u_3)$), we get:

>>>>

\begin{prop}
\label{prop:locanalytic}
\todo{Write a proposition.}
\end{prop}

@@

Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have important 
applications to effective computations with $p$-adic numbers or power 
series, that we are going to discuss now. Let us recall first that a 
datum involving elements of a complete ultrametric field $K$ (\emph{e.g.} 
$p$-adic numbers or power series) cannot be stored entirely on a computed 
because they basically convey an infinite amount of informations. That is 
why, instead of this datum itself, we generally work with neighborhoods 
of it. Such neighborhoods are typically sets of the form $x + H$ where 
$x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice in 
$E$. A basic example is the approximate $5$-adic number $1 + O(5^{10})$ 
which corresponds to the neighborhood of $1$ in $\Z_5$, which is the set 
of all $5$-adic numbers congruent to $1$ modulo $5^{10}$. In the context
of computer science, we will often use the notation $x + O(H)$ to refer 
to $x + H$.

Now, let us consider a function {\tt f} (in the sense of computer 
science) that takes as input an (approximate) element lying in an open 
subset $U$ of a $K$-Banach space $E$ and outputs another (approximate) 
element lying in an open subset $V$ of another $K$-Banach space $F$. Of 
course, this function can be modelised by a \emph{continuous} 
mathematical function $f : U \to V$. The relation is the following: when 
{\tt f} is called on the input $x + O(H)$, it outputs $x' + O(H')$ with 
$f(x+H) \subset x' + H'$. We say that {\tt f} does not generate 
artificial loss of precision if the above inclusion is an equality; it is 
often not the case as shown in \S \ref{ssec:stepbystep}.

\smallskip

Let us assume now that $f$ is locally analytic on $U$ and that $x$ and 
$H$ are chosen so that Lemma \ref{lem:main} applies. Then $f(x+H) = f(x) 
+ f'(x)(H)$ and the smallest possible $H'$ is $f'(x)(H)$. In other 
words, the function {\tt f} generates artificial loss of precision if 
the precision $H'$ it outputs is strictly larger that $f'(x)(H)$.
The aim of this section is to explain how, under the above hypothesis, 
one can implement the function {\tt f} so that it always outputs the
optimal precision $O(H')$ with $H' = f'(x)(H)$.

\subsubsection*{Forward computation}

The execution of the function {\tt f} yields a factorization:
$$f = f_n \circ f_{n-1} \circ \cdots \circ f_1$$
where the $f_i$'s correspond to each individual basic step (like 
addition, multiplication or affectation of variables); it is then a
``nice'' (in particular locally analytic) function. For all 
$i$, let $U_i$ denote the codomain of $f_i$. Concretely, $U_i$ is the 
set of all possible values of all variables which are defined in the 
program after the execution of $i$-th step. Mathematically, it is an 
open subset in some $K$-Banach space $E_i$. We have $U_n = V$ and the 
domain of $f_i$ is $U_{i-1}$ where, by convention, we have set $U_0 = 
U$.
For all $i$, we set $g_i = f_i \circ \cdots \circ f_1$ and $x_i = 
g_i(x)$.

When we execute the function {\tt f} on the input $x + O(H)$, we apply 
first $f_1$ to this input obtaining this way a first result $x_1 + 
O(H_1)$ and then go on with $f_2, \ldots, f_n$. At each step, we obtain 
a new intermediate result that we denote by $x_i + O(H_i)$. A way to 
guarantee that no artificial loss of precision are generated is then to 
ensure $H_i = f'_i(x)(H_{i-1}) = g_i'(x)(H)$ at each step. This can be 
achieved by reimplementing all primitives (addition, multiplication, 
\emph{etc.}) and make them compute at the same time the function $f_i$
they implement together with its differential and apply the latter to
the ``current'' lattice $H_i$.

There is nevertheless an important issue with this approach: in order to 
be sure that Lemma \ref{lem:main} applies, we need \emph{a priori} to 
compute the exact values of all $x_i$'s, which is of course not possible! 
Assuming that $g'_i(x)$ is surjective for all $i$, we can fix it as 
follows. For each $i$, we fix a lattice $\tilde H_i \subset H$ such that 
Lemma \ref{lem:main} applies to $g_i$, $x$ and $\tilde H_i$. Under our 
assumption, such lattices always exist and can be computed dynamically 
using Proposition \ref{prop:locanalytic}. Now, the equality $g_i(x + 
\tilde H_i) = x_i + g'_i(x)(\tilde H_i)$ means that any perturbation of 
$x_i$ by an element in $g'_i(x) (\tilde H_i)$ is induced by a 
perturbation of $x$ by an element in $\tilde H_i \subset H$. Hence, we 
can freely compute $x_i$ modulo $g'_i(x) (\tilde H_i)$ without changing 
the final result. Since $g'_i(x)(\tilde H_i)$ is a lattice in $E_i$, this 
remark makes possible the computation of $x_i$.

\begin{rmk}
In some cases, it would be possible to determine suitable lattices 
$\tilde H_i$ together with their images under $g'_i(x)$ (or, at
least, good approximations of them) before starting the computation 
by using mathematical arguments. If possible, this generally helps a
lot. We shall present in \S \ref{ssec:SOMOS-solution} an example of 
this.
\end{rmk}

\subsubsection*{Backward computation}

The previous approach works only if the $g'_i(x)$'s are all surjective. 
Unfortunately, this assumption is in general not fulfilled. Indeed, 
remember that the dimension of $E_i$ is roughly the number of used 
variables after the step $i$. It all $g_i'(x)$ were surjective, this 
would mean that the function {\tt f} never initializes a new variable!
In what follows, we propose another solution that does not assume the
surjectivity of $g'_i(x)$.

For $i \in \{1, \ldots, n\}$, define $h_i = f_n \circ \cdots \circ 
f_{i+1}$, so that we have $f = h_i \circ g_i$. On differentials, we 
have $f'(x) = h_i'(x_i) \circ g'_i(x)$. Since $f'(x)$ is surjective (by 
assumption), we deduce that all $h'_i(x_i)$ are surjective. Let $H'_i$ 
be a lattice in $E_i$ such that:
\begin{enumerate}[(a)] 
\item \label{item:Hi1}
$H'_i$ is contained in $H_i + \ker h'_i(x_i) = h'_i(x_i)^{-1}
\big(f'(x)(H)\big)$;
\item \label{item:Hi2}
Lemma \ref{lem:main} applies with the function $h_i$, the point
$x_i$ and the lattice $H'_i$.
\end{enumerate}
Applying Lemma \ref{lem:main}, we then get:
$$h_i(x_i + H'_i) = x_n + h'_i(x_i)(H'_i) \subset x_n + f'(x)(H).$$
Therefore, if after the $i$-th step of the execution of the function
{\tt f}, we modify the intermediate value $x_i$ by an element of
$H'_i$, the final result remains unchanged. In other words, it is
enough to compute $x_i$ modulo $H'_i$.

It is nevertheless not obvious to implement these ideas in practice
because when we enter in the $i$-th step of the execution of {\tt f},
we have not computed $h_i$ yet and hence are \emph{a priori} not able
to determine a lattice $H'_i$ satisfying the axioms \eqref{item:Hi1}
and \eqref{item:Hi2} above.
A possible solution to tackle this problem is to proceed in several
stages as follows: 
\begin{enumerate}[(1)]
\item \label{item:smallprec}
for $i$ going from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
at small precision (but enough for the second step) together with an
approximation of the Newton polygon of the higher order derivatives;
\item \label{item:determineHi}
for $i$ going from $n$ to $1$, we compute $h'_i(x_i)$ and
determines a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
\eqref{item:Hi2};
\item \label{item:finalcomp}
for $i$ going from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
and finally outputs $x_n + O\big(f'(x)(H)\big)$.
\end{enumerate}
Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
\cite{}), we can reuse in Step~\eqref{item:finalcomp} the computations 
already performed in Step~\eqref{item:smallprec}. The ``backward'' method 
we have just presented is then probably not much more expansive than the 
``forward'' method, although it is certainly much more difficult to 
implement.

We conclude this Section by remarking that the ``backward'' method seems 
to be particularly well suited to computations with lazy $p$-adic. Indeed 
in this setting, a target precision (\emph{i.e.} a precision on the 
output) is fixed and the software determines automatically the precision 
it needs on the input to perform the computation. To do this, it first 
builds the ``skeleton'' of the computation (\emph{i.e.} it determines the 
functions $f_i$ and eventually computes the $x_i$ at small precision when 
branching points occur and it needs to decide which branch it follows) 
and then runs over this skeleton in the reverse direction in order to 
determine (an upper bound of) the needed precision at each step. The 
similarity with our approach is noticeably strong.

@@

In this Section, we illustrate the theory developed in this paper by 
giving examples and applications. In \S \ref{ssec:differentials}, we 
compute the differential of a bunch of usual operations involving 
$p$-adic objects of different nature. Very often, we will observe that 
this computation is easy even if the underlying operation is 
sophisticated (\emph{e.g.} Gr\"obner basis).
In \S \ref{ssec:SOMOS-solution}, we study the SOMOS 4 sequence 
introduced in \S \ref{ssec:stepbystep} and, making a crucial use of 
Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic}, we design
a quite \emph{stable} algorithm for computing it.

@@

Another usual operations on polynomials is the factorization. Suppose 
that we are given a monic polynomial $P \in K[X]$ written as a product 
$P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
lemma implies that there exists a small neighborhood $\mathcal U$ of 
$P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
defined on the subset of $\mathcal U$ consisting of monic polynomials.
We want to differentiate $f$ at $P_0$. For this, we differentiate the 
equality $P = A B$ around $P_0$, obtaining 
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

@@

Other important tools related to matrices are usual factorizations as, 
for example, LU factorization. In what follows, let us agree to define a 
LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
Such a writing exists and is unique provided that all principal minors 
of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
defined over the Zariski-open set of matrices satisfying the above 
condition. In order to differentiate it, we differentiate the relation 
$M = LU$ and rewrite the result as follows:
$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
We remark that in the right hand side of the above formula, the first
summand is lower triangular and all its diagonal entries vanish whereas 
the second summand is upper triangular. Hence in order to compute $dL$
and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
left and $U$ on the right respectively. 
The above discussion extends almost \emph{verbatim} to LUP 
factorization; the only difference is that LUP factorizations are not 
unique but they are on a small neighborhood of $M$ if we fix the matrix 
$P$.

@@

Let us now consider the open subset of $M_{n,n-d}(K)$ consisting of 
matrices of full rank. The left kernel defines a mapping $\text{LK} : 
M_{n,n-d}(K) \to \Grass(d,n)$. Let us prove that it is differentiable 
and compute its differential around some point $M \in M_{n,n-d}(K)$. Of 
course, there exists a neighborhood $\mathcal U$ of $M$ whose image is 
entirely contained in a given chart $U_P$. We assume for simplicity that 
$P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
our chart to the application that takes $M$ to the unique matrix $N \in
M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 
= 0$. The implicit function theorem then implies that $\text{LK}$ is
differentiable. Furthermore, its differential satisfies the relation
$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
from which we can compute $dN$ by projecting on the $(n-d)$ last
columns.

@@

The lesson from this study is the following. If we have to compute the 
kernel of a matrix $M$ over $K$ known with finite precision, \emph{a 
priori} we first need to determine what is the rank of $M$. However, the 
rank function is not continuous --- it is only lower semi-continuous --- 
and hence cannot be determined from sure from $M$ (expect if $M$ has 
full rank). We then need to make an extra hypothesis and the more natural 
one is arguably to assume that $M$ has actually the minimal rank that it 
can have. With this convention, for instance, the kernel of the zero 
matrix $M \in M_n(K)$ \emph{known with finite precision} is always the 
whole space although there exist invertible matrices in any 
neighborhood of $M$.

@@

\ref{prop:locanalytic} to $\varphi_{i-1}$ and $\varphi_i$ (at the point 
$(u_0, u_1, u_2, u_3)$), we get:
